---
title: "gam-community-model"
author: "jt-miller"
date: "2025-04-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Use a General Additive Model (GAM) to model spring community flowering trends in Desert

Load Packages 
```{r}
library(data.table) # for fast read and write
library(tidyverse) # for data wrangling 
library(lubridate) # for date handling
library(mgcv) # for gam modeling
library(gratia) # for gam visuals
library(DHARMa) # for glm assumption checking
library(knitr) # for constructing result tables
library(gridExtra) # extra grid fxns for tables
library(grid) # more table functionality
```

#### Step 1. Data Processing

Load relevant data sources
```{r}
# read in the desert flowering annotated dataset
flower_data <- data.table::fread("/blue/guralnick/millerjared/SuperBlooms/data/processed/desert-observations-hexed.csv")
# load background RG iNaturalist observation data for vascular plants of these regions 
obs_data <- fread("/blue/guralnick/millerjared/SuperBlooms/data/processed/tracheophyte-desert-inats.csv")

# set up file pathing to gather rds objects of premade grid cell maps 
hex_map_files <- list.files("/blue/guralnick/millerjared/SuperBlooms/data/processed", pattern = "\\.rds$", full.names = TRUE)
```
Wrangle data
```{r}
# remove redundant annotated flowering data for unique observational records
flower_data <- dplyr::distinct(flower_data, obs_url, .keep_all = TRUE)

# standardize temporal fields for background obs data
obs_data <- obs_data %>% 
  mutate(eventDateParsed = parse_date_time(eventDate, orders = c("ymd_HMS", "ymd_HM", "ymd"))) %>% 
  mutate(doy = lubridate::yday(eventDateParsed)) %>% 
  mutate(year = lubridate::year(eventDateParsed)) %>% 
  mutate(month = lubridate::month(eventDateParsed))

# create temporal bins for flowering data
flower_data_temporalbins <- flower_data %>% # weekly bins
  dplyr::mutate(week_bin = cut(doy, breaks = seq(1, 366, by = 7), right = FALSE, labels = FALSE)) %>% # monthly bins
  dplyr::mutate(month_bin = cut(doy, breaks = seq(1,366, by = 30), right = FALSE, labels = FALSE)) %>% 
  dplyr::filter(!is.na(week_bin)) %>% # remove records that failed to be binned appropriately 
  dplyr::filter(!is.na(month_bin))

# do the same temporal binning for the background obs data
obs_data_temporalbins <- obs_data %>% # weekly bins
  dplyr::mutate(week_bin = cut(doy, breaks = seq(1, 366, by = 7), right = FALSE, labels = FALSE)) %>% # monthly bins
  dplyr::mutate(month_bin = cut(doy, breaks = seq(1,366, by = 30), right = FALSE, labels = FALSE)) %>% # remove records that failed to be binned appropriately 
  dplyr::filter(!is.na(week_bin)) %>% 
  dplyr::filter(!is.na(month_bin))

```


From an ecologist perspective, we want the data with the fine grain temporal and spatial resolution, however from knowing these data making a model to the finest scale (5km^2, day of year) will be challening to fit. Therefore, lets try the middle road of week bins and 50km^2 hex cells.
```{r}
# set up spatial & temporal bin retrieval vectors
hex_map_res <- c(100, 10, 25, 50, 5)
temporal_bins <- c("doy", "week_bin", "month_bin")

# grab relevant data for 50x50km & weekbins 
hex_grab <- 4
hex_cell_size <- hex_map_res[hex_grab]
temporal_name <- temporal_bins[2]

# create groupings for subsequent data summaries 
hex_name <- paste0("hex", hex_cell_size, "_id")
group_col <- list(sym("year"), sym(hex_name), sym(temporal_name))

# summarize into interpretable tables grouped hex cell-year-weekbin
hex_flower_data_bin <- flower_data_temporalbins %>% 
  group_by(!!!group_col) %>% # !!! splice operator
  summarize( 
    num_observations_flowering = n(), 
    uniqueFloweringSpecies = n_distinct(taxon_name),
    .groups = "drop")

# do the same for the the background obs data 
hex_obs_data_bin <- obs_data_temporalbins %>% 
  group_by(!!!group_col) %>% # use !!! splice operator     
  summarize(
    num_observations_total = n(), 
    uniqueObsSpecies = n_distinct(verbatimScientificName), 
    .groups = "drop"
  )

# merge flowering data with obs data 
hex_flower_obs_bin <- merge(hex_flower_data_bin, hex_obs_data_bin, by = c("year", hex_name, temporal_name), all.x = TRUE)

# due to our proxy of background data, we need to remove cases that do not make sense. throw these out for simplicity.
hex_flower_obs_bin <- hex_flower_obs_bin %>%       
  filter(!is.na(num_observations_total)) %>% # remove na obs
  mutate(obsFlowerRatioCheck = ifelse(num_observations_flowering <= num_observations_total, TRUE, FALSE)) %>% 
  filter(obsFlowerRatioCheck == TRUE)

# Our hexcells are not quite the same area contents at the edges of the desert due to shape clipping, we can create an area variable and put it into our model to better deal with this. 
hex_grid <- read_rds(paste0(hex_map_files[hex_grab]))
hex_info <- select(hex_grid, hex_name, area) %>% sf::st_drop_geometry()
hex_flower_obs_bin <- merge(hex_info, hex_flower_obs_bin , by = hex_name)
hex_flower_obs_bin <- hex_flower_obs_bin %>% 
  mutate(hex50_id = as.factor(hex50_id), 
         log_area = log(as.numeric(area) + 1))
```
#### 2. Data Removal
This data is messy by nature. We're going to remove some things that are obviously going to be an issue

1. Lets remove years that have low amounts of sampling, as these are unlikely to be informative to the model as a whole. 
```{r}
year_obs_check <- hex_flower_obs_bin %>% group_by(year) %>% summarize(n = sum(num_observations_flowering)) %>% arrange(desc(n))
year_obs_500 <- year_obs_check %>% filter(n >= 500) # set an arbitrary req
hex_flower_obs_bin <- hex_flower_obs_bin %>% filter(year %in% year_obs_500$year) %>% filter(year != 2024) # also remove year 2024, as this data is truncated by the 12th week
```

2. Lets remove data past week 22 marking the end of the month of May. This decision is because we're not looking at Fall bloom events (an equally interesting question in deserts...)
```{r}
hex_flower_obs_bin_spring <- hex_flower_obs_bin %>% 
  filter(week_bin <= 22)
```

#### Step 3. Apply GAM & Check Model Assumptions
```{r}
mod_nb <- gam(num_observations_flowering ~ # Y response var
                s(year, bs = "cs") + # overall yearly trend
                s(week_bin, bs = "cs") + # seasonal effect
                s(hex50_id, bs = "re") + # code hexcell id as a random effect
                s(log_area, bs = "tp") + # log of area with a thin-plate spline 
                num_observations_total, # use number of background obs as a fixed effect in order to account for sampling effort. 
              method = "REML", # method recommended by Gavin Simpsons ecological forecasting course. 
              data = hex_flower_obs_bin_spring, 
              family = nb()) # use family negative binomial, if dispersion is low this will act like a poisson anyway.

summary(mod_nb)
```

Check GAM & GLM assumptions to see if we're in the clear
```{r}
# check whether basis search did well enough. 
gam.check(mod_nb)

# check what the smooths looked like 
draw(mod_nb)
ggsave("/blue/guralnick/millerjared/SuperBlooms/figures/gam-model-trends.png", width = 10, height = 10)
# draw a rootagram to infer how well the data matches expected area of fit 
rootogram(mod_nb) %>% draw()
ggsave("/blue/guralnick/millerjared/SuperBlooms/figures/gam-model-rootogram.png", width = 10, height = 10)

# use the DHARMa package to assess how well we fullfill typical GLM assumptions
# check residuals fit 
png("/blue/guralnick/millerjared/SuperBlooms/figures/gam-model-residual-simu.png", width = 600, height = 480)
sim_res <- DHARMa::simulateResiduals(mod_nb, plot = TRUE)
dev.off()

# check for crazy outliers
png("/blue/guralnick/millerjared/SuperBlooms/figures/gam-model-residual-simu-outliers.png", width = 600, height = 480)
sim_test_outliers <- testOutliers(sim_res, type = "bootstrap")
dev.off()
# check for overdispersion, especially if the test above is sig
png("/blue/guralnick/millerjared/SuperBlooms/figures/gam-model-residual-simu-dispersion.png", width = 600, height = 480)
testDispersion(sim_res)
dev.off()
```
Use positive residuals that sit outside of 95% of the data as a way to pull out candidate superblooms
```{r}
# use a residuals test to pull out and label candidate superblooms
hex_flower_obs_bin_spring$resid <- residuals(mod_nb, type = "pearson")
hex_flower_obs_bin_spring$fitted <- fitted(mod_nb)

# ID large pos residuals (z-score > 2)
hex_flower_obs_bin_spring$superBloomCandidate <- hex_flower_obs_bin_spring$resid > 2

# fitler down to only include these superbloom candidate bins
possibleSuperblooms <- filter(hex_flower_obs_bin_spring, superBloomCandidate == TRUE)
possibleSuperblooms %>% group_by(year) %>% summarize(n = n()) %>% arrange(desc(n))

top10_superbloom_years <- hex_flower_obs_bin_spring %>%
  filter(superBloomCandidate == TRUE) %>%
  group_by(year) %>%
  summarize(n = n()) %>%
  arrange(desc(n)) %>%
  slice_head(n = 5)

# Display table
kable(top10_superbloom_years, 
      col.names = c("Year", "Number of Superbloom Bins"), 
      caption = "Top 5 Years with Most Superbloom Candidate Bins")

# create a output png table version

# Create the table grob
tbl <- tableGrob(top10_superbloom_years, rows = NULL)

# Create the title grob (multi-line)
title <- textGrob("Top 5 Years with Most Superbloom \n Candidate Bins",
                  gp = gpar(fontsize = 12, fontface = "bold"))

# Calculate actual title height
title_height <- grobHeight(title) + unit(1, "mm")  # Add a little buffer

# Combine title and table
tbl_with_title <- arrangeGrob(
  title, tbl,
  ncol = 1,
  heights = unit.c(title_height, unit(1, "npc") - title_height)
)

# Save to PNG
png("/blue/guralnick/millerjared/SuperBlooms/figures/top5_superblooms.png", width = 200, height = 175)
grid.draw(tbl_with_title)
dev.off()


top5_flowering_obs <- hex_flower_obs_bin_spring %>%
  group_by(year) %>%
  summarize(n = n()) %>%
  arrange(desc(n)) %>%
  slice_head(n = 5)

# Display table
kable(top5_flowering_obs, 
      col.names = c("Year", "Number of Flowering Bins"), 
      caption = "Top 5 Years with Flowering Bins")

# create output table png again
# Create the table grob
tbl <- tableGrob(top5_flowering_obs, rows = NULL)

# Create the title grob (multi-line)
title <- textGrob("Top 5 Years with Most Flowering \n Observation Bins",
                  gp = gpar(fontsize = 12, fontface = "bold"))

# Calculate actual title height
title_height <- grobHeight(title) + unit(1, "mm")  # Add a little buffer

# Combine title and table
tbl_with_title <- arrangeGrob(
  title, tbl,
  ncol = 1,
  heights = unit.c(title_height, unit(1, "npc") - title_height)
)

# Save to PNG
png("/blue/guralnick/millerjared/SuperBlooms/figures/top5_flower_obs_bins.png", width = 200, height = 175)
grid.draw(tbl_with_title)
dev.off()

```

Visual To assist 
```{r}
# Create a standard normal distribution
x <- seq(-4, 4, length.out = 1000)
y <- dnorm(x)

# Create a data frame
df <- data.frame(x = x, y = y)

# Create the plot
ggplot(df, aes(x, y)) +
  geom_line(size = 1) +
  geom_area(data = subset(df, x > 2), aes(x, y), fill = "red", alpha = 0.6) +
  labs(title = "Standard Normal Distribution",
       subtitle = "Residuals > 2 (superbloom candidates)",
       x = "Pearson Residual",
       y = "Density") +
  theme_minimal(base_size = 14)

ggsave("/blue/guralnick/millerjared/SuperBlooms/figures/pearsons-res-vis.png", width = 10, height = 10)
``` 

Now use these data to create a species extraction table and some figures.
```{r}
# extract species data 
# load data
flower_data <- fread("/blue/guralnick/millerjared/SuperBlooms/data/processed/desert-observations-hexed.csv")
superbloom_candidates <- possibleSuperblooms # the data for superblooms

# we only want unique observations (currently the data is structured so that multiple images of one obs event take up uniq rows)
flower_data <- dplyr::distinct(flower_data, obs_url, .keep_all = TRUE)

# create temporal bins 
flower_data_temporalbins <- flower_data %>% 
  # weekly bins
  dplyr::mutate(week_bin = cut(doy, breaks = seq(1, 366, by = 7), right = FALSE, labels = FALSE)) %>%
  # monthly bins
  dplyr::mutate(month_bin = cut(doy, breaks = seq(1,366, by = 30), right = FALSE, labels = FALSE)) %>% 
  # remove tail end data 
  dplyr::filter(!is.na(week_bin)) %>% 
  dplyr::filter(!is.na(month_bin))

# create summary tables comparing the species composition between candidate superblooms 
select_flowering_data <- flower_data_temporalbins %>% 
  dplyr::select(year, family, taxon_name, hex50_id, week_bin) %>% 
  mutate(hex50_id = as.factor(hex50_id))

select_flowering_superbloomCandidateData <- select_flowering_data %>% 
  inner_join(superbloom_candidates, by = c("year", "hex50_id", "week_bin"))

uniq_cells_num <- length(unique(select_flowering_superbloomCandidateData$hex50_id))
superbloom_spatial_summary <- select_flowering_superbloomCandidateData %>% 
  group_by(hex50_id, taxon_name) %>%
  dplyr::summarize(n = n()) %>%
  ungroup() %>% 
  group_by(taxon_name) %>% 
  dplyr::summarize(distinctCellsOccupied = n_distinct(hex50_id), 
                   percentOccupied = distinctCellsOccupied/uniq_cells_num)
uniq_wkbin_num <- length(unique(select_flowering_superbloomCandidateData$week_bin))
superbloom_temporal_summary <- select_flowering_superbloomCandidateData %>% 
  group_by(week_bin, taxon_name) %>%
  dplyr::summarize(n = n()) %>%
  ungroup() %>% 
  group_by(taxon_name) %>% 
  dplyr::summarize(distinctWkBinsOccupied = n_distinct(week_bin), 
                   percentOccupied = distinctWkBinsOccupied/uniq_wkbin_num)

```
Create tables for representing these summaries
```{r}
top5_superbloom_spatial_summary <- superbloom_spatial_summary %>%
  arrange(desc(distinctCellsOccupied)) %>%
  slice_head(n = 5)

# Display table
kable(top5_superbloom_spatial_summary, 
      col.names = c("Taxon", "distinct cells occupied", "percent of total cells occupied"), 
      caption = "Top 5 spatial taxa for superbloom candidates")

top5_superbloom_temporal_summary <- superbloom_temporal_summary %>%
  arrange(desc(distinctWkBinsOccupied)) %>%
  slice_head(n = 5)

# Display table
kable(top5_superbloom_temporal_summary, 
      col.names = c("Taxon", "distinct week bins occupied", "percent of total week bins occupied"), 
      caption = "Top 5 temporal bins taxa for superbloom candidates")
```
Save those as pngs 
```{r}
# Rename columns for prettier headers
top5_superbloom_spatial_summary_renamed <- top5_superbloom_spatial_summary %>%
  rename(
    Taxon = taxon_name,
    `Cells Occupied` = distinctCellsOccupied,
    `% Total Cells Occupied` = percentOccupied
  )

# Create the table grob
tbl_spatial <- tableGrob(top5_superbloom_spatial_summary_renamed, rows = NULL)

# Create the title grob
title_spatial <- textGrob("Top 5 Spatial Taxa for Superbloom Candidates",
                           gp = gpar(fontsize = 12, fontface = "bold"))

# Calculate dynamic height of title
title_height_spatial <- grobHeight(title_spatial) + unit(2, "mm")

# Combine and render
tbl_spatial_with_title <- arrangeGrob(
  title_spatial, tbl_spatial,
  ncol = 1,
  heights = unit.c(title_height_spatial, unit(1, "npc") - title_height_spatial)
)

# Save to PNG
png("/blue/guralnick/millerjared/SuperBlooms/figures/top5_superbloom_spatial_summary.png", width = 400, height = 175)
grid.draw(tbl_spatial_with_title)
dev.off()

# Rename columns for prettier headers
top5_superbloom_temporal_summary_renamed <- top5_superbloom_temporal_summary %>%
  rename(
    Taxon = taxon_name,
    `Week Bins Occupied` = distinctWkBinsOccupied,
    `% Week Bins Occupied` = percentOccupied
  )

# Create the table grob
tbl_temporal <- tableGrob(top5_superbloom_temporal_summary_renamed, rows = NULL)

# Create the title grob
title_temporal <- textGrob("Top 5 Temporal Taxa for Superbloom Candidates",
                           gp = gpar(fontsize = 12, fontface = "bold"))

# Calculate dynamic height of title
title_height_temporal <- grobHeight(title_temporal) + unit(2, "mm")

# Combine and render
tbl_temporal_with_title <- arrangeGrob(
  title_temporal, tbl_temporal,
  ncol = 1,
  heights = unit.c(title_height_temporal, unit(1, "npc") - title_height_temporal)
)

# Save to PNG
png("/blue/guralnick/millerjared/SuperBlooms/figures/top5_superbloom_temporal_summary.png", width = 400, height = 175)
grid.draw(tbl_temporal_with_title)
dev.off()

# === Prepare Spatial Table ===
top5_superbloom_spatial_summary_renamed <- top5_superbloom_spatial_summary %>%
  rename(
    Taxon = taxon_name,
    `Cells Occupied` = distinctCellsOccupied,
    `% Total Cells Occupied` = percentOccupied
  )

tbl_spatial <- tableGrob(top5_superbloom_spatial_summary_renamed, rows = NULL)
title_spatial <- textGrob("Top 5 Spatial Taxa for Superbloom Candidates",
                          gp = gpar(fontsize = 12, fontface = "bold"))
title_height_spatial <- grobHeight(title_spatial) + unit(2, "mm")

tbl_spatial_with_title <- arrangeGrob(
  title_spatial, tbl_spatial,
  ncol = 1,
  heights = unit.c(title_height_spatial, unit(1, "npc") - title_height_spatial)
)

# === Prepare Temporal Table ===
top5_superbloom_temporal_summary_renamed <- top5_superbloom_temporal_summary %>%
  rename(
    Taxon = taxon_name,
    `Week Bins Occupied` = distinctWkBinsOccupied,
    `% Week Bins Occupied` = percentOccupied
  )

tbl_temporal <- tableGrob(top5_superbloom_temporal_summary_renamed, rows = NULL)
title_temporal <- textGrob("Top 5 Temporal Taxa for Superbloom Candidates",
                           gp = gpar(fontsize = 12, fontface = "bold"))
title_height_temporal <- grobHeight(title_temporal) + unit(2, "mm")

tbl_temporal_with_title <- arrangeGrob(
  title_temporal, tbl_temporal,
  ncol = 1,
  heights = unit.c(title_height_temporal, unit(1, "npc") - title_height_temporal)
)

# === Combine Both Tables Vertically with Consistent Size ===
combined_tables <- arrangeGrob(
  tbl_spatial_with_title,
  tbl_temporal_with_title,
  ncol = 1,
  heights = unit.c(unit(1, "npc") * 0.5, unit(1, "npc") * 0.5)  # Each gets 50%
)

# === Save Final Combined Table ===
png("/blue/guralnick/millerjared/SuperBlooms/figures/top5_spatial_temporal_combined.png", width = 500, height = 350)
grid.draw(combined_tables)
dev.off()
```
# Create a jaccard similarity matrix
```{r}
#### try some matrix comparisons:

# create an event id by temporal
select_flowering_superbloomCandidateData  <- select_flowering_superbloomCandidateData  %>% 
  mutate(event_id = paste(year, week_bin, sep = "_"))

# create a field of lists of taxa per event
taxa_list <- select_flowering_superbloomCandidateData %>% 
  distinct(taxon_name, event_id) %>% 
  group_by(event_id) %>% 
  summarize(taxa = list(unique(taxon_name)), .groups = "drop")

# do pairwise combinations of events
event_pairs <- expand.grid(event_id1 = taxa_list$event_id, 
                           event_id2 = taxa_list$event_id, 
                           stringsAsFactors = FALSE)
# calc the shared taxa among events
compare_fxn <- function(event_id1, event_id2, taxa_list){
  taxa1 <- taxa_list$taxa[taxa_list$event_id == event_id1][[1]]
  taxa2 <- taxa_list$taxa[taxa_list$event_id == event_id2][[1]]
  intersect_taxa <- intersect(taxa1, taxa2)
  
  tibble(
    event_id1 = event_id1, 
    event_id2 = event_id2, 
    n_taxa_1 = length(taxa1), 
    n_taxa_2 = length(taxa2), 
    n_shared = length(intersect_taxa), 
    pct_shared_1 = length(intersect_taxa)/length(taxa1),
    pct_shared_2 = length(intersect_taxa)/length(taxa2)
  )
}

comp_tb <- purrr::pmap_dfr(
  event_pairs,
  .f = compare_fxn,
  taxa_list = taxa_list
)

# create a jaccard similairy index 
comp_tb <- comp_tb %>% 
  mutate(jaccard = n_shared / (n_taxa_1 + n_taxa_2 - n_shared))

ggplot(comp_tb, aes(x = event_id1, y = event_id2, fill = jaccard)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(option = "plasma", name = "Jaccard Similarity") +
  labs(
    title = "Pairwise Taxonomic Similarity Across Superbloom Events",
    x = "Event ID",
    y = "Event ID"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    panel.grid = element_blank()
  )

ggsave("/blue/guralnick/millerjared/SuperBlooms/figures/jaccard-similarity-superbloomEvents.png", width = 14, height = 14)


```


