---
title: "simulate-superblooms"
author: "jt-miller"
date: "2025-06-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---
title: "detect-superbloom-simu"
author: "jt-miller"
date: "2025-04-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Try and create the most basic hierarchical model of flowering abundance 
Assumptions <br>
- Each year $y$ (for y = 1, ...,30) flowering follows a unimodal phenology curve <br>
- The number of flowering individuals on day-of-year (DOY) $d$ is normally distributed around a year specific phenological mean  $\mu_y$, with year-specific standard deviation $\sigma_y$ which can represent the duration of flowering(?). <br>
- The total flowering abundance per year may vary, modeled as a latent year-specific abundance term $N_y$ <br>

Let, <br>
- $A_{y,d}$ be the abundance of flowering indivduals on day $d$ of year $y$ 
- $N_y$ be the total flowering abudance in year $y$
- $\mu_y$ be the mean flowering date for year $y$ 
- $\sigma_y$ be the duration of flowering period for year $y$ (?)

Math to form a Gaussian phenology curve per year:
$$
A_{y,d} = N_y \times NormalPDF(d|\mu_y,\sigma_y) \\ 
where, \\ 
NormalPDF(d|\mu,\sigma) = \frac{1}{\sqrt2{\pi\sigma}} exp(-\frac{1}{2\sigma^2}(d - \mu)^2)
$$
Create a simu of this basic model, note that I'll be using hyperparameters, because peak flowering day (mu_y) and flowering duration (sigma_y) vary across years but are not completely independent. They are samples from higher level (hyper) distributions. These hyperparameters describe the distributions *from which year-specific parameters are drawn*. 

E.g. If most years have peak flowering around doy 100, then we model: 
$$
\mu_y \sim Normal(\mu_\mu, \sigma_\mu)
$$
where $\mu_\mu$ is the long term average phenological mean, and $\sigma_\mu$ reflects interannual variability. These define the prior distributions for year-level parameters. 
```{r}
library(ggplot2)
library(dplyr)
 
# define simu parms
n_years <- 30
days <- 1:365

# set up hyperparms 
mu_mu <- 100 # average flowering peak across years
sigma_mu <- 10 # variability in peak flowering phenology across years
mu_sigma <- 15 # spread/duration of flowering
sigma_sigma <- 5 # spread/duration of flowering 
mu_logN <- 9 # log total abundance 
sigma_logN <- 0.5 # log sigma 

# Simu yearly parms 
mu_y <- rnorm(n_years, mean = mu_mu, sd = sigma_mu) # draw each year's flowering mean mu_y for Normal Distributions defined by the hyperparameters mu_mu & sigma_mu
sigma_y <- abs(rnorm(n_years, mean = mu_sigma, sd = sigma_sigma)) # draw each year's standard deviation sigma_y from Normal distributions defined by the hyperparameters mu_sigma and sigma_sigma
N_y <- rlnorm(n_years, meanlog = mu_logN, sdlog = sigma_logN) # simulate N_y using a log-normal distribution


# Simulate flowering abundance curves per year
sim.data <- do.call(rbind, lapply(1:n_years, function(y) {
  doy_density <- dnorm(days, mean = mu_y[y], sd = sigma_y[y])
  A_yd <- N_y[y] * doy_density
  
  data.frame(
    year = y, 
    doy = days, 
    abundance = A_yd
  )
}))

# Visualize a subset of years
sim.data %>% 
  filter(year %in% c(1,5,10,20, 30)) %>% 
  ggplot(aes(x = doy, y = abundance, color = factor(year))) +
  geom_line() +
  labs(title = "Simu Flowering Abudundance Curves", 
       x = "Day of Year", y = "Abundance", color = "Year") + 
    theme_minimal()
```
Currently N_y is constructed is such a way that it just randomly draws froma log-normal distribution. This means that we are 'randomly' experiencing high abundance flowering years vs low abundance flowering years. <br>

### We can expand this model (without yet adding biology) by additional latent superbloom indicator parameter S_y. 
$$
S_y \sim Bernoulli(p) \\
log(N_y) = \alpha_0 + \alpha_1 \times S_y \\
A_{y,d} = N_y \times NormalPDF(d |\mu_y, \sigma_y)
$$
Here, we're assuming that there are two different regimes of flowering. Normal years where S_y = 0, such that log(N_y) = alpha_0 resulting in moderate/low abundance of flowering. Superbloom years where S_y = 1, such that log(N_y) = alpha_0 + alpha_1 resulting in high abundance of flowering. This is a two component mixture model on N_y where S_y chooses which distribution to draw from and alpha_1 defines how extreme superblooms are. <br>

We're going to stay in the hyperparameter context, so that we can apply inference methods later. Here, lets define some events <br> 

Let $p_{bloom} \sim Beta(2,8)$ so that each year is sampled like $S_y \sim Bernoulli(p_{bloom})$, so that the expected proportion of superbloom years is 20%
$$
E[p_{bloom}] = \frac{2}{2 + 8} = 0.20 
$$
```{r}
# set up parms 
n_years <- 30
days <- 1:365

# Hyperparms
mu_mu <- rnorm(1, mean = 100, sd = 10) # mean DOY peak across years
sigma_mu <- abs(rnorm(1, mean = 0, sd = 5)) # variability across years

mu_sigma <- rnorm(1, mean = 12, sd = 3) # spread/duration of flowering 
sigma_sigma <- abs(rnorm(1, mean = 0, sd = 3)) # spread/duration of flowering

alpha_0 <- 9 # means that N_y ~ exp(9) in normal years
alpha_1 <- 2.5 # means that N_y ~ exp(9 + 2.5) in superbloom years

p_bloom <- rbeta(1, shape1 = 2, shape2 = 8) # beta distribution, the conjugate prior for the Bernoulli and Binomial, should center over 10% probability of superbloom

# year specific parameters
mu_y <- rnorm(n_years, mean = mu_mu, sd = sigma_mu) # draw each year's flowering mean mu_y for Normal Distributions defined by the hyperparameters mu_mu & sigma_mu
sigma_y <- abs(rnorm(n_years, mean = mu_sigma, sd = sigma_sigma)) # draw each year's standard deviation sigma_y from Normal distributions defined by the hyperparameters mu_sigma and sigma_sigma
S_y <- rbinom(n_years, size = 1, prob = p_bloom) # draw from a binomial distribution for whether this should be a superbloom year, given the probability of the superbloom. 

log_N_y <- alpha_0 + alpha_1 * S_y # Determine log-abundance of flowering
N_y <- exp(log_N_y) # Exponentiate

# Simulate flowering abundances
simu.data <- lapply(1:n_years, function(y) {
  doy <- days
  curve <- dnorm(doy, mean = mu_y[y], sd = sigma_y[y])
  abundance <- N_y[y] * curve
  data.frame(year = y, doy = doy, abundance = abundance, S_y = S_y[y])
}) %>% bind_rows()

# Normalize data for visualizing 
simu.data <- simu.data %>% 
  group_by(year) %>% 
  mutate(abundance_scaled = abundance/max(abundance)) %>% 
  ungroup()

# Visualize
set.seed(1)
select_years <- sample(1:n_years, 15) # take some random set of years

# Visualize a subset of years
simu.data %>% 
  filter(year %in% c(1,2,3,4,5,6,7,8,9,10)) %>% 
  ggplot(aes(x = doy, y = abundance, color = factor(year), linetype = factor(S_y))) +
  geom_line() +
  scale_linetype_manual(values = c("solid", "dashed"), 
                        labels = c("Normal", "Superbloom"), 
                        name = "Bloom type") +
  labs(title = "Simu Flowering Abudundance Curves", 
       x = "Day of Year", y = "Abundance", color = "Year") + 
    theme_minimal()

```
Now lets start simple, fit to a 2nd degree polynomial model per year
```{r}
library(tidyverse)

simu.data_count <- simu.data %>%
  mutate(count = rpois(n(), lambda = abundance)) %>%  # simulate count data from abundance
  group_by(year, doy, S_y) %>%
  summarise(count = sum(count), .groups = "drop")
# prep doy 
# Fit a quadratic model per year
poly_fit <- function(df) {
  tryCatch({
    lm(count ~ poly(doy, degree = 2, raw = TRUE), data = df)
  }, error = function(e) NULL)
}

model_fits_poly <- simu.data_count %>%
  group_by(year) %>%
  nest() %>%
  mutate(model = map(data, poly_fit))
```

Compare
```{r}
library(broom)


model_preds <- model_fits_poly %>%
  mutate(predicted = map2(model, data, ~ {
    if (is.null(.x)) return(NULL)
    augment(.x, newdata = .y)
  })) %>%
  unnest(predicted)

years_to_plot <- 1:9

model_preds %>%
  filter(year %in% years_to_plot) %>%
  ggplot(aes(x = doy)) +
  geom_col(aes(y = count), fill = "grey70") +  # histogram bars
  geom_line(aes(y = .fitted, color = factor(S_y)), size = 1.2) +  # model fit
  facet_wrap(~ year, scales = "free_y") +
  scale_color_manual(values = c("0" = "blue", "1" = "red"), 
                     labels = c("Normal", "Superbloom"),
                     name = "Bloom Type") +
  labs(title = "Quadratic Fit on Flowering Count Data per Year",
       x = "Day of Year", y = "Count") +
  theme_minimal()

```
```{r}
library(tidyverse)

set.seed(1)
n_years <- 30
days <- 1:365

# --- Hyperparameters ---
mu_mu <- rnorm(1, mean = 100, sd = 10) 
sigma_mu <- abs(rnorm(1, mean = 0, sd = 5)) 
mu_sigma <- rnorm(1, mean = 12, sd = 3)
sigma_sigma <- abs(rnorm(1, mean = 0, sd = 3))
alpha_0 <- 9
alpha_1 <- 2.5
p_bloom <- rbeta(1, 2, 8)

# --- Year-specific parameters ---
mu_y <- rnorm(n_years, mu_mu, sigma_mu)
sigma_y <- abs(rnorm(n_years, mu_sigma, sigma_sigma))
S_y <- rbinom(n_years, 1, p_bloom)
log_N_y <- alpha_0 + alpha_1 * S_y
N_y <- exp(log_N_y)

# --- Simulate flowering counts ---
simu.data <- map_dfr(1:n_years, function(y) {
  lambda_doy <- N_y[y] * dnorm(days, mean = mu_y[y], sd = sigma_y[y])
  counts <- rpois(length(days), lambda = lambda_doy)
  tibble(year = y, doy = days, count = counts, S_y = S_y[y])
})

ggplot(simu.data %>% filter(year %in% 1:9), aes(x = doy, y = count)) +
  geom_col() +
  facet_wrap(~ year, scales = "free_y") +
  theme_minimal()
```
Try fitting the model
```{r}
# fit a test year
year_data <- simu.data %>% filter(year == 1)
x <- glm(count ~ poly(doy, 2, raw = TRUE), family = poisson, data = year_data)
# Predict the fitted values
year_data$fitted <- predict(x, type = "response")

# Create a new data frame for the trend line
trend_line <- data.frame(
  doy = seq(min(year_data$doy), max(year_data$doy), by = 1),
  fitted = predict(x, newdata = data.frame(doy = seq(min(year_data$doy), max(year_data$doy), by = 1)), type = "response")
)

# Create the plot
ggplot(year_data, aes(x = doy, y = count)) +
  geom_col() +
  geom_line(data = trend_line, aes(x = doy, y = fitted), color = "red") +
  labs(title = "Count Data by Day of Year with Fitted Trend Line",
       subtitle = "Poisson Regression Model with Quadratic Term",
       x = "Day of Year",
       y = "Count") +
  theme_classic()
```



### Adjusting N_y to be influenced by key covariates thought to cause superblooms
Here, Im adding covariates that will modify N_y such that we have some underyling ecology that triggers such events. 
